\thetitle

### 2008-05-28 paper.tex
Corrected a few typos & errors.

Despite great breakthroughs in understanding and methodology, the networking community remains without a ``cookbook'' for generating realistic \caps{LAN} workloads. By ``cookbook,'' we mean an explicit method for generating network traffic---down to the size and transmission time of each application data unit. This lack is particularly relevant to wireless research, since the medium fundamentally resource-limited, and consequently local traffic patterns can have a huge impact on overall network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.

@@

Paxson and Floyd observed in~\cite{Paxson95} that the interplay between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback. Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06:dissertation}. This presents a fundamental difficulty, however, in that it presumes that we know the intent of endpoints: what \textit{would} they have done under different conditions? While we can speculate about what an individual node might hypothetically do, we currently hardly understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations. Especially in the wireless setting, a fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback. Accordingly, in this paper we attempt to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without multi-level feedback. It is important to realize that while this does not provide a final picture, we currently lack even a first-order understanding of the effect of different workloads on performance. This first-order understanding is an essential initial step.

There have been a significant number of studies of large wireless network deployments~\cite{Tang99,Balachandran02,Balazinska03,Kotz02,Henderson04,Schwab04,Chinchilla04,Jardosh05:ewind}. These analyses have described a wide variety of aspects of wireless network behavior, and provide much insight into the workings of real, deployed wireless networks. These studies present a broad analysis of general system features and trends of specific corporate wireless local-area networks (\caps{WLAN}s)~\cite{Tang99,Balazinska03}, university campus \caps{WLAN}s~\cite{Tang00,Kotz02,Chinchilla04,Schwab04,Henderson04,Tuduce05}, and temporary \caps{WLAN}s at conference venues~\cite{Balachandran02,Jardosh05:ewind}. They also provide a large body of raw data for subsequent analysis and modeling research. Our work provides the methodology for turning this rich foundation of field data into usable, realistic models of workload for a wide variety of networking situations.

@@

The {packets-bytes-duration} representation is clearly overly simplistic and very incomplete. It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time and flow size unspecified; it says nothing about the source or destination of the flow. At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets, etc. into a vector space. The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows. We take a highly detailed but intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the flow type, source node, destination node, start time, flow size, and empirical distributions of packet sizes and inter-packet intervals. If there are $f$ flows in the entire workload pattern, then the collective behavior can be described with six matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order. We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
\begin{tabbing}
\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
$D\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
\>\bullet\> type: \>$T$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$,\\
\>\bullet\> sources: \>$S$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$,\\
\>\bullet\> destinations: \>$D$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$,\\
\>\bullet\> start times: \>$A$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_t}$,\\
\>\bullet\> sizes in bytes: \>$B$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$,\\
\>\bullet\> packet sizes: \>$Z$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
\>\bullet\> inter-packet intervals: \>$V$\>$=\parensx{\vec{v}_i}$\>$=\parensx{v_{ij}}$\>$\in\R^{f \times d_v}$.
\end{tabbing}
The total behavior is the direct product of these components: $F = \seq{T,S,D,A,B,Z,V} \in \R^{f \times \Sigma}$ where $\Sigma=256+2n+d_t+d_b+d_z+d_v$. Many common traffic models can be expressed succinctly and precisely in terms of transformations or conditions on this behavior matrix and its constituents. In the following sections, we explain in detail how each flow's behavior is represented in these matrices. Finally, we describe how a model instance stochastically generates network traffic.

>>>>

Despite great breakthroughs in understanding and methodology, the networking community remains without a ``cookbook'' for generating realistic \caps{LAN} workloads. By ``cookbook,'' we mean an explicit method for generating network traffic---down to the size and transmission time of each application data unit. This lack is particularly relevant to wireless research, since the medium is fundamentally resource-limited, and consequently local traffic patterns can have a huge impact on overall network performance~\cite{Karpinski07:realism,Karpinski07:cbr-failure}.

@@

Paxson and Floyd observed~\cite{Paxson95} that the interplay between endpoint behavior and the network conditions is inherently \textit{closed-loop} in the sense that it is potentially affected by complex feedback. Traffic models typically attempt to preserve the closed-loop behavior of network traffic~\cite{Avallone04,Hernandez06:dissertation}. This presents a fundamental difficulty, however, in that it presumes that we know the intent of endpoints: what \textit{would} they have done under different conditions? While we can speculate about what an individual node might hypothetically do, we currently hardly understand the impact of the full-network traffic pattern upon performance metrics at all---even without trying to account for hypothetical reactions to alternate situations. Especially in the wireless setting, a fuller understanding of total network behavior must be reached before we can sensibly tackle the complexities of multi-level behavioral feedback. Accordingly, in this paper we attempt to provide a first-order approximation of complete network behavior by studying the response of performance metrics to \textit{open-loop} traffic models without multi-level feedback. It is important to realize that while this does not provide a final picture, we currently lack even a first-order understanding of the effect of different workloads on performance. This first-order understanding is an essential initial step.

There have been many studies of large wireless network deployments~\cite{Tang99,Balachandran02,Balazinska03,Kotz02,Henderson04,Schwab04,Chinchilla04,Jardosh05:ewind}. These analyses have described a wide variety of aspects of wireless network behavior, and provide much insight into the workings of real, deployed wireless networks. These studies present a broad analysis of general system features and trends of specific corporate wireless local-area networks (\caps{WLAN}s)~\cite{Tang99,Balazinska03}, university campus \caps{WLAN}s~\cite{Tang00,Kotz02,Chinchilla04,Schwab04,Henderson04,Tuduce05}, and temporary \caps{WLAN}s at conference venues~\cite{Balachandran02,Jardosh05:ewind}. They also provide a large body of raw data for subsequent analysis and modeling research. Our work provides the methodology for turning this rich foundation of field data into usable, realistic models of workload for a wide variety of networking situations.

@@

The {packets-bytes-duration} representation is clearly overly simplistic and very incomplete. It captures nothing about variability of packet sizes, nor about the ``burstiness'' of packet transmissions; it leaves start time and flow size unspecified; it says nothing about the source or destination of the flow. At the opposite extreme, it is possible to embed the exact start time of each flow, and the exact size and transmission times of its packets, etc. into a vector space. The difficulty comes in doing so in a manner that makes the summation of vectors correspond naturally and meaningfully to the aggregate behavior of flows. We take a highly detailed but intermediate approach, suggested by the results of~\cite{Karpinski07:realism}: to reproduce network performance accurately, it suffices to know, for each flow, the flow type, source node, destination node, start time, flow size, and empirical distributions of packet sizes and inter-packet intervals. If there are $f$ flows in the entire workload pattern, then the collective behavior can be described with six matrices; the rows of each matrix are the behavior vectors of the flows, indexed in some chosen order. We use the following notation for the behavior matrices and their constituent vectors:\vspace{-0.25em}
\begin{tabbing}
\hspace{1em}\=\bullet\hspace{0.5em}\=\hspace{9.75em}\=
$D\:$\=$=\parensx{\vec{d}_i}\:$\=$=\parensx{d_{ij}}\:$\=\kill
\>\bullet\> type: \>$T$\>$=\parensx{\vec{t}_i}$\>$=\parensx{t_{ij}}$\>$\in\R^{f \times 255}$,\\
\>\bullet\> sources: \>$S$\>$=\parensx{\vec{s}_i}$\>$=\parensx{s_{ij}}$\>$\in\R^{f \times n}$,\\
\>\bullet\> destinations: \>$D$\>$=\parensx{\vec{d}_i}$\>$=\parensx{d_{ij}}$\>$\in\R^{f \times n}$,\\
\>\bullet\> start times: \>$A$\>$=\parensx{\vec{a}_i}$\>$=\parensx{a_{ij}}$\>$\in\R^{f \times d_t}$,\\
\>\bullet\> sizes in bytes: \>$B$\>$=\parensx{\vec{b}_i}$\>$=\parensx{b_{ij}}$\>$\in\R^{f \times d_b}$,\\
\>\bullet\> packet sizes: \>$Z$\>$=\parensx{\vec{z}_i}$\>$=\parensx{z_{ij}}$\>$\in\R^{f \times d_z}$,\\
\>\bullet\> inter-packet intervals: \>$V$\>$=\parensx{\vec{v}_i}$\>$=\parensx{v_{ij}}$\>$\in\R^{f \times d_v}$.
\end{tabbing}
The total behavior is the horizontal concatenation of these component matrices:
\begin{align}
F = \seq{T,S,D,A,B,Z,V} \in \R^{f \times \Sigma}
\end{align}
where $\Sigma=255+2n+d_t+d_b+d_z+d_v$. Many common traffic models can be expressed succinctly and precisely in terms of transformations or conditions on this behavior matrix and its constituents. In the following sections, we explain in detail how each flow's behavior is represented in these matrices. Finally, we describe how a model instance stochastically generates network traffic.
